Neighborhoods and Felony Disenfranchisement: The Case of New York City

### 2019-05-21 markdown/kmorris_thesis.Rmd
clean up written part

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "April 22, 2019"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "April 22, 2019" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->

@@

Felony disenfranchisement has received much attention over the past few years. This is true in both the academic world (where researchers have dug into the political ramifications in such races as the 2000 presidential election) and in the political world as states such as Florida and Louisiana have gradually moved to dismantle their systems of disenfranchisement. Throughout this process, however, many researchers have focused on the *individual* effects of felony disenfranchisement. They have examined the effect of disenfranchisement on the political participation of individuals, including after they are no longer disenfranchised, but have generally neglected to examine the intersection of spatially concentrated policing and incarceration patterns with the political effects of disenfranchisement. This paper seeks to increase our understanding of the spatial implications of felony disenfranchisement by examining the character of disenfranchisement in New York City.

This paper also seeks to complicate the narratives surrounding the end of felony disenfranchisement. Although advocates are correct to push for an end to disenfranchisement, a change to the political system may not be sufficient to reincorporate the voices of formerly incarcerated individuals into the democratic process. In addition to laying out the spatial implications of felony disenfranchisement in two major American cities, this paper also explores the impact of two different policies intended to dismantle the disenfranchisement program. Specifically, I investigate the effect of Executive Order 181 (which ended the disenfranchisement of parolees in New York State) and  Amendment 4 (which restored the voting rights of Floridians who had completed all terms of their sentences). The effects of these policies make clear that, while the end of felony disenfranchisement is necessary, active steps must be taken to reincorporate the formerly disenfranchised.

@@

In all but two states (Maine and Vermont), felony disenfranchisement laws ensure that American citizens convicted of felony offenses lose the right to vote for at least some period of time. In some states, such as Oregon and Massachusetts, individuals lose that right only for the period in which they are actively incarcerated. In other states, notably Kentucky and Iowa, felony convictions result in lifelong disenfranchisement unless a returned citizen receives an individual pardon from the state's governor [@bcj_laws]. This variation in laws flows directly from language in the Fourteenth Amendment which allows states to revoke individuals' voting rights "for participation in rebellion, or other crime." The definition of "other crime," left so vague in the Constitution, is now generally used by states to encompass any felony offense at all. The Supreme Court, in cases such as *Richardson v. Ramirez*, has upheld states' right to do just that. Collectively, these laws disenfranchise as many as `r 6.1 - 1.4` million American citizens. Of these, the majority are no longer incarcerated, but are living and working in their communities [@sentencing_2016].^[The figures reported in @sentencing_2016 have been adjusted to reflect the impact of Amendment 4 in Florida.]

In any discussion of felony disenfranchisement in the United States, it is imperative to acknowledge the central role played by race and white supremacy. As Traci Burch has explained, "If policies restricting the voting rights of offenders disparately affect one racial group or party, it is because such policies were intended to" [@Burch2010]. The historical record is undeniable. Previous research established that the presence of nonwhite potential voters is associated with the implementation of felony disenfranchisement policies and that these policies were often adopted during Jim Crow to reduce the political power of Black Americans [@Behrens2003]. In Florida felony disenfranchisement was added to the state constitution in 1868. Afterwards, a lawmaker boasted that the effort had been made in order to prevent Florida from being "niggerized" [@florida_1868].

The racial imbalance of felony disenfranchisement laws are not confined to the 19th century. Although the Voting Rights Act of 1965 did much to improve access to the ballot box for minorities, it did nothing to undermine the explicitly racialized system of disenfranchisement. Indeed, as the United States has vastly increased the reach of the carceral system in the post-Civil Rights era, the implications of felony disenfranchisement have only grown. As of 2016, more than 10 percent of African Americans were disenfranchised in 9 states. In Kentucky, the state with the highest level of disenfranchised Black residents, more than one in four Black adults are barred from casting a ballot. Although Black adults made up just `r percent(29932674 / 247219588, accuracy = 0.1)` of the voting age population in 2016, they accounted for `r percent(2228118 / 6106327, accuracy = 0.1)` of the disenfranchised population [@sentencing_2016].

@@

In the aftermath of the 2000 presidential election, academic interest in the political implications of felony disenfranchisement was stirred thanks to a paper from Christopher Uggen and Jeff Manza [-@Uggen2002]. George W. Bush's margin of victory in Florida in 2000 was famously just 537 votes. In their 2002 paper, Uggen and Manza estimate the likely partisan composition of the disenfranchised population with felony convictions in their past. They estimate that, if this group had been allowed to vote, they would have supported Al Gore. Their enfranchisement, Uggen and Manza posit, would have tipped the presidential contest and resulted in the election of Al Gore. They based their estimates on the voting patterns of *non*-disenfranchised individuals who were demographically similar to the disenfranchised population. Though much of the research conducted since their 2002 study has undermined some of their key assumptions (namely, that formerly incarcerated individuals turnout and cast ballots similarly to those who have not been incarcerated), Uggen and Manza convincingly demonstrated that felony disenfranchisement can have material political consequences. In the years after the Uggen and Manza published their paper, scholars sought to investigate the effect of felony disenfranchisement on African American turnout [@Miles2004], young adult turnout [@Hjalmarsson2010], and other questions. Some of this research compared states and regions with differing disenfranchisement regimes to estimate these effects [@Miles2004; @Ochs2006]. Others have used survey data or interviews to construct their estimates [@Uggen2004; @Drucker2005].

In a series of papers in between 2009 and 2011, researchers developed methods for directly estimating the turnout of formerly disenfranchised individuals. Haselswerdt [-@Haselswerdt2009] matched release data and voter registration data from Erie County, NY, to estimate turnout among a small group of formerly incarcerated individuals. Traci Burch [-@Burch2010; -@Burch2011] expanded upon this matching methodology to estimate the voting patterns of formerly disenfranchised individuals in a range of states. She uses release data from states' Departments of Corrections and their Registered Voter Files to estimate the number of formerly incarcerated individuals who went on to register to vote. Using the Registered Voter Files, she is also able to estimate the party affiliation of formerly incarcerated individuals (in states with party registration) and the turnout rate among these individuals. Her methodology has been used to investigate other questions surrounding the voting patterns of formerly incarcerated individuals under different circumstances and to examine the impact of changes in disenfranchisement policy [@Meredith2013; @Meredith2015].

A number of papers have also explored the impact that felony disenfranchisement policies have on turnout among non-disenfranchised residents. King and Erickson [-@King2016], for instance, leverage state-level variation in disenfranchisement laws to estimate the impact that felony disenfranchisement has on turnout among African Americans. They use data from the 2004 Current Population Survey Voting and Registration Supplement to determine individual-level turnout. They include estimates of the share of African Americans who are disenfranchised in each state from @locked_out to explore the impact of these policies on eligible voters. Ultimately, they argue that "African American disenfranchisement plays a unique role in predicting African American voter turnout." They conclude that disenfranchisement has spillover effects for African American voters: where more African Americans are disenfranchised, elibible African Americans are less likely to cast a ballot. These findings are in line with other research that has explored whether the effects of disenfranchisement extend beyond those whose voting rights are directly suspended [@Bowers2009; @Ochs2006; @Walker2014]. As @Bowers2009 sum up: "[I]t is not solely the direct vote of ex-felons that is denied through these laws. [Felony disenfranchisement] impacts the political power of communities that extends beyond felons' collateral penalty."

Although scholars have established that felony disenfranchisement decreases turnout among Black voters at the *state* level, relatively little research has been done on how felony disenfranchisement operates at the sub-state level. Though we know that Black voters are generally less likely to cast a ballot when they live in a state with strict disenfranchisement laws, less work has been done exploring the types of neighborhoods in which these effects are concentrated, and the impact these laws might have at the local level. @Burch2013 is an exception to this. In this paper, Burch explores the depressive effect of disenfranchisement laws at the local level in North Carolina by examining census block-group level turnout and involvement with the criminal justice system. She determines that "at high concentrations, imprisonment and community supervision have an unequivocally demobilizing effect of neighborhoods." This paper seeks to expand on her work by replicating her findings in New York City and by introducing a new way of measuring the impact.

# Spatial Effects of Concentrated Policing and Incarceration {-}

XX

@@

This paper begins by examining the effect of felony disenfranchisement in the New York City Mayoral election of 2017. I am primarily interested in whether felony disenfranchisement has impacts on local political distribution. As discussed above, policing and incarceration patterns have historically targeted communities of color, wreaking havok on the social fabric of these neighborhoods. It is possible, however, that these effects are even broader than the direct focus on policing has acknowledged. We know that felony disenfranchisement systematically removes voters from certain neighborhoods, but it is not clear whether enough voters are removed relative to the electorate to meaningfully distort neighborhood representation. However, to the extend that felony disenfranchisement has a depressive effect on *eligible* voters, it is possible that disenfranchisement policies are powerful enough to materially reduce the representatition of certain parts of the city. Although turnout is not a perfect proxy for political representation, there is some evidence that policians are more responsive to groups that turnout at higher rates. @Martin2013, for instance, documents the relationship between voter turnout and the responsiveness of Congressional representatives. "[H]igher citizen participation is rewarded," they argue, "with enhanced policy responsiveness." If the neighborhoods most impacted by felony disenfranchisement turn out at lower rates, they may find that their elected representatives are less likely to support their needs. @Griffin2005 find similar effects in the United States Senate, reporting that "voter preferences predict the aggregate roll-call behavior of Senators while nonvoter preferences do not." If felony disenfranchisement has a depressive effect on turnout, it is likely that it is responsible for politicians being less responsive to the particular preferences of voters in neighborhoods where voters are removed --- some of the most disadvantaged neighborhoods to begin with.

After examining the effect of felony disenfranchisement on voter turnout at the neighborhood level, I consider the efficacy of a program intended to undo the impact of felony disenfranchisement. In April of 2018, Governor Andrew Cuomo signed Executive Order 181 which restored voting rights to New Yorkers on parole.^[Prior to Executive Order 181, residents convicted of felonies in prison and on parole were barred from voting, while individuals sentenced to probation did not have their voting rights revoked. Today, only those in prison for felony offenses are disenfranchised.] Parole officers were required to provide registration forms the parolees under their supervision and inform them of their voting rights. In theory, such a policy could increase turnout and registration for parolees. To test whether the policy was successful, I compare the turnout rates in the 2018 general election of three groups who finished parole in 2018: those who finished parole immediately before the policy change, those who finished parole immediately after the policy change, and those who finished parole immediately after the election. These three groups allow me to explore whether the re-extension of voting rights while on parole has an impact on turnout (by comparing the first two groups) and whether being under formal supervision while being allowed to vote results in higher turnout than being allowed to vote but being under supervision (by comparing the third group to the first two).

The first part of this project seeks to refine how we think about felony disenfranchisement in the United States. Rather thank think about it purely through the prism of those directly impacted (residents who are actively disenfranchised) or through a prism of race (by examining the impact of disenfranchisement on African-American turnout at the state level), I seek to introduce the importance of physical space into the conversation. This will allow us to both better understand the mechanisms through which the depressive effects operate and also to understand more deeply where these effects are concentrated. Of course, understanding the impact of felony disenfranchisement is not enough: we must also be testing whether our efforts to undo its legacy are effective. The second part of this project seeks to do just that.

@@

Although understanding whether felony disenfranchisement has impacts on turnout among different social groups (namely, racial minorities) is of great importance, understanding the explicitly spatial dimension of the problem is also an important task. Many political decisions bear on the spatial distribution of resources (XX); funding for parks or schools, for instance, can impact all members of a neighborhood, regardless of their race or their income. As such, even a resident who is not directly affected by felony disenfranchisement policies may lose political power thanks to these policies; if many of my neighbors cannot vote, I am joined by fewer voters in advocating for neighborhood-specific investment and services. This is, of course, tied up closely with race; New York City remains highly racially segregated, and therefore disentangling neighborhood effects from race effects becomes difficult. Nevertheless, these effects cross race lines in integrated neighborhoods.

Despite dire over-incarceration in some neighborhoods, the number of incarcerated individuals is relatively low compared to the number of voters. In New York, for instance, `r comma(readRDS("./temp/number_incarcerated.rds"))` individuals were imprisoned in New York State in early 2019, compared with `r round(readRDS("./temp/active_voters_0319.rds")/1000000, digits = 1)` million actively registered voters. Despite the low share of residents who are directly disenfranchised, there is reason to believe the policy impacts more individuals than just those imprisoned. As discussed above, previous research has demonstrated that felony disenfranchisement reduces turnout even among Black voters whose rights are not stripped. This research has found, in particular, that eligible Black voters are less likely to cast a ballot in states where felony disenfranchisement policies are harsher, an effect often referred to as *de facto* disenfranchisement. The existence of such *de facto* disenfranchisement is bolstered by a large body of empirical work on the social nature of voting. Many scholars have established that being in a community where voting is common makes individuals more likely to vote. Children whose parents vote, for instance, are more likely to vote as adults [XX]; XX add more examples here. Many of these relationships are spatially bounded: family members are likely to live in the same home, or on the same block. Literature from urban sociology has established that social networks are largely spatially bounded. To the extent that felony disenfranchisement policies have depressive effects on turnout in the social and filial networks of the imprisoned, these policies are likely to be closely concentrated in the neighborhoods where the disenfranchised live.

Although research on the impact of local turnout on city-wide policy is scarce, @Anzia2019 examines the impact of senior turnout on "senior-friendly" policy at the city level. Anzia does not find that senior turnout in general increases the likelihood of senior-friendly policies, but that elected officials *are* responsive to senior turnout when seniors "are a more cohesive, meaningful group." American cities are highly segregated by race and by class, but less so by age. As such, Anzia's study does not speak directly to the impact of *neighborhood* turnout rates on city policy. An extension of her paper, however, implies that what she finds might be probitive to the effect of neighborhood turnout rates. Insofar as neighborhoods vote as a bloc, their turnout rates may influence city policy.

@@

@@

The City of New York also makes the location of all NYPD arrests since 2006 available publicly. This data includes the latitude and longitude of the arrest, the date of the offense, and the class of the alleged offense. This analysis uses only arrests for felony charges.

@@

The state does not make a unified database of parolees whose voting rights have been restored available to the public. However, the Department of Corrections and Community Supervision website's Parolee Lookup includes a flag indicating whether someone's voting rights haave been restored. By using identification numbers from the data obtained from the state to lookup parolees on the website, I was able to construct a list of the individuals who were on parole and had their rights restored.^[This list was compiled by using a webscraper written in the Python language.]

## Voter File Data {-}
Most states in the United States are required to maintain files with information on all registered voters in the state. In New York, this information is publicly available from the Board of Elections. It includes information on all registered voters, including: first and last name; date of birth; home address; vote history; and other information. The New York State Voter File also includes information on voters who were previously registered but have since been purged, either because they moved, died, or were incarcerated for a felony offense. Because the voter file includes information on voters who have been purged because of a felony conviction, we are able to reconstruct the vote history of individuals who are no longer registered because of their conviction.

## Geocoding {-}
Voters' home addresses were converted to latitudes and longitudes using a geocoder provided by SmartyStreets. These latitudes and longitudes were then converted to census block groups, census tracts, and city council districts using the statistical program R.

@@

## Identification of Lost Voters {-}
In this analysis, I offer a slightly different defintion of "lost voter" than much of the literature. Many recent papers have attempted to identify relationships between the number of *potentially* lost voters and turnout. Here, I explore whether the disenfranchisement of voters with a history of participating in elections is related to neighborhood turnout. In this analysis, "lost voters" are individuals ineligible to cast a ballot on a given election day who have cast a ballot in the previous ten years. To study the effect of felony disenfranchisement on voting at the local level (and its potential implications for the distribution of political power at the local level), I use turnout in the most recent election for city-wide office in New York --- the mayoral election which took place on November 7^th^, 2017. Lost voters, therefore, are all individuals who were incarcerated or on parole on November 7^th^, 2017, and had cast a ballot between 2007 and 2016.

These individuals are identified by matching Department of Corrections and Community Supervisions data to the voter file using the methodology developed by Burch [-@Burch2010; -@Burch2011]. I use first, middle, and last names, and dates of birth, to join these two datasets.^[Matching on first and last names and dates of birth can result in false positive matches, especially in large states like New York. Following @Meredith2013, I report estimates of the potential impact of these errors in [Appendix B]] Figure \ref{fig:citymap} shows where these lost voters lived before going to prison, with city council districts also included. There were `r comma(readRDS("./temp/lost_count_nyc.rds"))` such lost voters within New York City as of the 2017 general election, and `r comma(nrow(readRDS("./temp/ids_of_lost_voters.rds")))` statewide.

```{r citywide-map, echo = FALSE, message = F, warning = F, fig.width = 5, fig.height = 5, fig.cap = "\\label{fig:citymap}Lost Voters on Election Day, 2017", out.extra = '', fig.pos = "H", fig.align = "center"}
j <- readRDS("./output/city_map.RDS")
print(j)
```
The spatial concentration of lost is readily apparent. In some communities, such as Greenwich Village and Brooklyn Heights, hardly any voters were disqualified from participating in the 2017 elections. In other communities, such as Harlem and Central Brooklyn, large numbers of individuals were not allowed to cast a ballot for mayor.

@@

Matching methodologies, of course, only allow us to test the effect of being treated --- here, losing a voter for the 2017 election. The model above does not allow for different effects on turnout based on *how many* voters a neighborhood lost. In Table \ref{tab:trad-reg} below, I adopt a standard ordinary least squares regression to investigate whether lost voters are associated with lower turnout rates in the 2017 election. This regression uses the same covariates that were used in the matching procedure described above. Each neighborhood is weighted by its voting age population, and robust standard errors are clustered by city council district.

>>>>

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
bibliography: "../research/citations.bib"
link-citations: yes
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->

@@

Felony disenfranchisement has received much attention over the past few years. This is true in both the academic world (where researchers have dug into the political ramifications in such races as the 2000 presidential election) and in the political world as states such as Florida and Louisiana have gradually moved to dismantle their systems of disenfranchisement. Many researchers, however, have focused on the effects of felony disenfranchisement on individuals, and on racial minorities as a group. They have examined the effect of disenfranchisement on the political participation of individuals, including after they are no longer disenfranchised, and the spillover effects disenfranchisement has on eligible voters of color. Researchers have generally neglected to examine the intersection of spatially concentrated policing and incarceration patterns with the political effects of disenfranchisement. This paper seeks to increase our understanding of the spatial implications of felony disenfranchisement by examining the relationship between disenfranchisement and neighborhood turnout in New York City.

This paper also seeks to complicate the narratives surrounding the end of felony disenfranchisement. Although advocates are correct to push for an end to disenfranchisement, a change to the political system may not be sufficient to reincorporate the voices of formerly incarcerated individuals into the democratic process. In addition to laying out the spatial implications of felony disenfranchisement in New York City, this paper also explores the impact of one policy intended to weaken the disenfranchisement program. Specifically, I investigate the effect of New York Governor Andrew Cuomo's Executive Order 181 which ended the disenfranchisement of parolees in New York State. The effect of these policies make clear that, while the end of felony disenfranchisement is necessary, it is not enough: Active steps must be taken to reincorporate the formerly disenfranchised.

@@

In all but two states (Maine and Vermont), felony disenfranchisement laws ensure that American citizens convicted of felony offenses lose the right to vote for at least some period of time. In some states, such as Oregon and Massachusetts, individuals lose that right only for the period in which they are actively incarcerated. In other states, notably Kentucky and Iowa, felony convictions result in lifelong disenfranchisement unless a returned citizen receives an individual pardon from the state's governor [@bcj_laws]. This variation in laws flows directly from language in the Fourteenth Amendment which allows states to revoke individuals' voting rights "for participation in rebellion, or other crime." The definition of "other crime," left so vague in the Constitution, is now generally used by states to disenfranchize citizens for any felony offense at all. The Supreme Court, in cases such as *Richardson v. Ramirez*, has upheld states' right to do just that. Collectively, these laws disenfranchise as many as `r 6.1 - 1.4` million American citizens. Of these, the majority are no longer incarcerated, but are living and working in their communities [@sentencing_2016].^[The figures reported in @sentencing_2016 have been adjusted to reflect the impact of Amendment 4 in Florida.]

In any discussion of felony disenfranchisement in the United States, it is imperative to acknowledge the central role played by race and white supremacy. As Traci Burch has explained, "If policies restricting the voting rights of offenders disparately affect one racial group or party, it is because such policies were intended to" [@Burch2010]. The historical record is undeniable. Previous research has established that the presence of nonwhite potential voters is associated with the implementation of felony disenfranchisement policies and that these policies were often adopted during Jim Crow to reduce the political power of Black Americans [@Behrens2003]. In Florida felony disenfranchisement was added to the state constitution in 1868. Afterwards, a lawmaker boasted that the effort had been made in order to prevent Florida from being "niggerized" [@florida_1868].

The racial imbalance of felony disenfranchisement laws are not confined to the 19th century. Although the Voting Rights Act of 1965 did much to improve access to the ballot box for minorities, it did nothing to undermine the explicitly racialized system of disenfranchisement. Indeed, as the United States has vastly increased the reach of the carceral system in the post-Civil Rights era, the implications of felony disenfranchisement have only grown. As of 2016, more than 10 percent of Black Americans were disenfranchised in 9 states. In Kentucky, the state with the highest level of disenfranchised Black residents, more than one in four Black adults are barred from casting a ballot. Although Black adults made up just `r percent(29932674 / 247219588, accuracy = 0.1)` of the voting age population in 2016, they accounted for `r percent(2228118 / 6106327, accuracy = 0.1)` of the disenfranchised population [@sentencing_2016].

Recent data Florida make the victims of felony disenfranchisement laws especially clear. Prior to January, 2019, Floridians convicted of felony offenses were permanently disenfranchised unless they were individually pardoned by a clemency board. In November, 2018, voters passed a ballot initiative amending the state constitution to end permanent disenfranchisement. In the first three months after permanent disenfranchisement was ended, 44 percent of formerly incarcerated Floridians who registered to vote were Black --- compared with just 13 percent of the statewide electorate [@Morris2019].

@@

In the aftermath of the 2000 presidential election, academic interest in the political implications of felony disenfranchisement was stirred thanks to a paper from Christopher Uggen and Jeff Manza [-@Uggen2002]. George W. Bush's margin of victory in Florida in 2000 was famously just 537 votes. In their 2002 paper, Uggen and Manza estimate the likely partisan composition of the disenfranchised population with felony convictions in their past. They estimate that, if this group had been allowed to vote, they would have supported Al Gore by a wide margin. Their enfranchisement, Uggen and Manza argued, would have tipped the presidential contest and resulted in the election of Al Gore. They based their estimates on the voting patterns of eligible individuals who were demographically similar to the disenfranchised population. Though much of the research conducted since their 2002 study has pushed back against some of their key assumptions (namely, that formerly incarcerated individuals turn out at the same rate as those who have not been incarcerated), Uggen and Manza convincingly demonstrated that felony disenfranchisement can have material political consequences. In the years after the Uggen and Manza published their paper, scholars sought to investigate the relationship between felony disenfranchisement and Black and youth turnout [@Miles2004; @Hjalmarsson2010]. Some of this research compared states and regions with differing disenfranchisement regimes to estimate these effects [@Miles2004; @Ochs2006]. Others have used survey data or interviews to construct their estimates [@Uggen2004; @Drucker2005].

In a series of papers in between 2009 and 2011, researchers developed methods for directly estimating the turnout of formerly disenfranchised individuals. Haselswerdt [-@Haselswerdt2009] matched release data and voter registration data from Erie County, NY, to estimate turnout among a small group of formerly incarcerated individuals. Traci Burch [-@Burch2010; -@Burch2011] expanded upon this matching methodology to estimate the voting patterns of formerly disenfranchised individuals in a range of states. She used release data from states' Departments of Corrections and their registered voter files to identify formerly incarcerated individuals who went on to register to vote. Using the registered voter files, she was also able to estimate the party affiliation of formerly incarcerated individuals (in states with party registration) and their turnout rates. Her methodology has been used to investigate other questions surrounding the voting patterns of formerly incarcerated individuals under different circumstances and to examine the impact of changes in disenfranchisement policy [@Meredith2013; @Meredith2015].

A number of papers have also explored the impact that felony disenfranchisement policies have on turnout among non-disenfranchised residents. King and Erickson [-@King2016], for instance, leverage state-level variation in disenfranchisement laws to estimate the impact that felony disenfranchisement has on turnout among Black Americans. They use data from the 2004 Current Population Survey Voting and Registration Supplement to determine individual-level turnout. They include estimates of the share of Black Americans who are disenfranchised in each state from @locked_out to explore the impact of these policies on eligible voters. Ultimately, they argue that "African American disenfranchisement plays a unique role in predicting African American voter turnout." They conclude that disenfranchisement has spillover effects for Black voters: where more Black residents are disenfranchised, elibible Black voters are less likely to cast a ballot. These findings are in line with other research that has explored whether the effects of disenfranchisement extend beyond those whose voting rights are directly suspended [@Bowers2009; @Ochs2006; @Walker2014]. As @Bowers2009 sum up: "[I]t is not solely the direct vote of ex-felons that is denied through these laws. [Felony disenfranchisement] impacts the political power of communities that extends beyond felons' collateral penalty."

Although scholars have established that felony disenfranchisement decreases turnout among Black voters at the *state* level, relatively little research has been done on how felony disenfranchisement operates at the sub-state level. Though we know that Black voters are generally less likely to cast a ballot when they live in a state with strict disenfranchisement laws, less work has been done exploring the impact these laws might have at the local level. @Burch2013 is an exception to this. In this paper, Burch explores the depressive effect of disenfranchisement laws at the local level in North Carolina by examining census block-group level turnout and involvement with the criminal justice system. Using a standard ordinary least squares regression determined that "at high concentrations, imprisonment and community supervision have an unequivocally demobilizing effect of neighborhoods." This paper seeks to expand on her work by replicating her findings in New York City; by using a different estimation technique; and by using a different definiton of "lost voters."

@@

This paper begins by exploring the effect of felony disenfranchisement on neighborhood turnout in the New York City Mayoral election of 2017. Policing and incarceration patterns have historically targeted communities of color, wreaking havok on the social fabric of these neighborhoods [XX]. It is possible, however, that these effects are even broader than the direct focus on policing has acknowledged. We know that felony disenfranchisement systematically removes voters from certain neighborhoods, but it is not clear whether enough voters are removed relative to the electorate to meaningfully distort neighborhood representation. To the extent that felony disenfranchisement has a depressive effect on *eligible* voters, it is possible that these policies are powerful enough to materially reduce the representatition of certain parts of the city. Although turnout is not a perfect proxy for political representation, there is some evidence that policians are more responsive to groups that turn out at higher rates. @Martin2013, for instance, documents the relationship between voter turnout and the responsiveness of Congressional representatives. "[H]igher citizen participation is rewarded," they argue, "with enhanced policy responsiveness." If the neighborhoods most impacted by felony disenfranchisement turn out at lower rates, they may find that their elected representatives are less likely to support their needs. @Griffin2005 find similar effects in the United States Senate, reporting that "voter preferences predict the aggregate roll-call behavior of Senators while nonvoter preferences do not."

Although research on the impact of local turnout on city-wide policy is scarce, @Anzia2019 examines the impact of senior turnout on "senior-friendly" policy at the city level. Anzia does not find that senior turnout in general increases the likelihood of senior-friendly policies, but that elected officials *are* responsive to senior turnout when seniors "are a more cohesive, meaningful group." American cities are highly segregated by race and by class, but less so by age. As such, Anzia's study does not speak directly to the impact of *neighborhood* turnout rates on city policy. An extension of her paper, however, implies that what she finds might be probitive to the effect of neighborhood turnout rates. Insofar as neighborhoods vote as a bloc, their turnout rates may influence city policy. If felony disenfranchisement has a depressive effect on turnout, it is likely that it is responsible for politicians being less responsive to the particular preferences of voters in neighborhoods where voters are removed --- some of the most neglected neighborhoods to begin with.

After examining the effect of felony disenfranchisement on voter turnout at the neighborhood level, I consider the efficacy of a program intended to undo one piece of the felony disenfranchisement apparatus. In April of 2018, Governor Andrew Cuomo signed Executive Order 181 which restored voting rights to New Yorkers on parole.^[Prior to Executive Order 181, residents convicted of felonies in prison and on parole were barred from voting, while individuals sentenced to probation did not have their voting rights revoked. Today, only those in prison for felony offenses are disenfranchised.] Parole officers were required to provide registration forms the parolees under their supervision and inform them of their voting rights. In theory, such a policy could increase turnout and registration for parolees. To test whether the policy was successful, I compare the turnout rates in the 2018 general election of three groups who finished parole in 2018: those who finished parole immediately before the policy change, those who finished parole immediately after the policy change, and those who finished parole immediately after the election. These three groups allow me to explore whether the re-extension of voting rights while on parole has an impact on turnout (by comparing the first two groups) and whether being under formal supervision while being allowed to vote results in higher turnout than being allowed to vote but being under supervision (by comparing the third group to the first two).

The first part of this project seeks to refine how we think about felony disenfranchisement in the United States. Rather thank think about it purely through the prism of those directly impacted (residents who are actively disenfranchised) or through a prism of race (by examining the impact of disenfranchisement on Black turnout at the state level), I seek to introduce the importance of physical space into the conversation. This will allow us to both better understand the mechanisms through which the depressive effects operate and also to understand more deeply where these effects are concentrated. Of course, understanding the impact of felony disenfranchisement is not enough: we must also test whether our efforts to undo its legacy are effective. The second part of this project seeks to do just that.

@@

Although understanding whether felony disenfranchisement impacts turnout among different social groups is of great importance, understanding the explicitly spatial dimension of the problem is also an important task. Many political decisions bear on the spatial distribution of resources (XX); funding for parks or schools, for instance, can impact all members of a neighborhood, regardless of their race, income, or criminal justice involvement. As such, even a resident who is not directly affected by felony disenfranchisement policies may lose political power thanks to these policies; if many of my neighbors cannot vote, I am joined by fewer voters in advocating for neighborhood-specific investment and services. This is, of course, tied up closely with race; New York City remains highly racially segregated. Thanks to spatially disparate policing patterns, any depressive effect is therefore likely to be concentrated among Black voters.

Despite dire over-incarceration in some neighborhoods, the number of incarcerated individuals is relatively low compared to the number of voters. In New York, for instance, `r comma(readRDS("./temp/number_incarcerated.rds"))` individuals were imprisoned in New York State in early 2019, compared with `r round(readRDS("./temp/active_voters_0319.rds")/1000000, digits = 1)` million actively registered voters. Despite the low share of residents who are directly disenfranchised, there is reason to believe the policy impacts more individuals than just those imprisoned. As discussed above, previous research has demonstrated that felony disenfranchisement reduces turnout even among Black voters whose rights are not stripped. This research has found, in particular, that eligible Black voters are less likely to cast a ballot in states where felony disenfranchisement policies are harsher, an effect often referred to as *de facto* disenfranchisement. The existence of such *de facto* disenfranchisement is bolstered by a large body of empirical work on the social nature of voting. Many scholars have established that being in a community where voting is common makes individuals more likely to vote. Children whose parents vote, for instance, are more likely to vote as adults [XX]; XX add more examples here. Literature from urban sociology has established that social networks are largely spatially bounded [XX]. To the extent that felony disenfranchisement policies have depressive effects on turnout in the social and filial networks of the imprisoned and paroled, these effects are likely to be closely concentrated in the neighborhoods where the disenfranchised live.

@@

* Super low turnout
* Immediate effects on future participation - Ariel's paper [@WHITE2019]
* Changes in Notification Laws (M&M)

@@

@@

The state does not make a unified database of parolees whose voting rights have been restored available to the public. However, the NYSDOCCS Parolee Lookup website includes a flag indicating whether someone's voting rights have been restored. By using identification numbers from the parolee data obtained from the state, I constructed a list of the individuals who were on parole and had their rights restored.^[This list was compiled by using a webscraper written in the Python language.]

## Voter File Data {-}
Most states in the United States are required to maintain files with information on all registered voters. In New York, this information is publicly available from the Board of Elections. It includes information on all registered voters, including: first and last name; date of birth; home address; vote history; and other information. The New York State Voter File also includes information on voters who were previously registered but have since been purged, either because they moved, died, or were incarcerated for a felony offense. The first section of this paper uses a snapshot of the registered voter file from April 30^th^, 2018. The second section of this paper uses a snapshot from March 3^rd^, 2019.

## Geocoding {-}
Voters' home addresses were converted to latitudes and longitudes using a geocoder provided by SmartyStreets. I then used the statistical software R to map these latitudes and longitudes to census block groups, census tracts, and city council districts using shapefiles publicly available from the Census Bureau and the City of New York.

@@

## Identification of Lost Voters {-}
In this analysis, I offer a slightly different defintion of "lost voter" than much of the literature. Many recent papers have attempted to identify relationships between the number of disenfranchised residents --- *potentially* lost voters --- and turnout. Here, I explore whether the disenfranchisement of residents with a history of participating in elections is related to neighborhood turnout. In this analysis, "lost voters" are individuals ineligible to cast a ballot on a given election day who have cast a ballot in the previous ten years. To study the effect of felony disenfranchisement on voting at the local level (and its potential implications for the distribution of political power at the local level), I use turnout in the most recent non-special election for city-wide office in New York --- the mayoral election which took place on November 7^th^, 2017. Lost voters, therefore, are all individuals who were incarcerated or on parole on November 7^th^, 2017, and had cast a ballot between 2007 and 2016.

These individuals are identified by matching NYSDOCCS data to the voter file. I use first, middle, and last names, and dates of birth, to join these two datasets.^[Matching on first and last names and dates of birth can result in false positive matches, especially in large states like New York. Following @Meredith2013, I discuss the potential impact of these errors in [Appendix A].] Figure \ref{fig:citymap} shows where these lost voters lived before going to prison, with city council districts also included. There were `r comma(readRDS("./temp/lost_count_nyc.rds"))` such lost voters within New York City as of the 2017 general election, and `r comma(nrow(readRDS("./temp/ids_of_lost_voters.rds")))` statewide.

```{r citywide-map, echo = FALSE, message = F, warning = F, fig.width = 5, fig.height = 5, fig.cap = "\\label{fig:citymap}Lost Voters on Election Day, 2017", out.extra = '', fig.pos = "H", fig.align = "center"}
j <- readRDS("./output/city_map.RDS")
print(j)
```
The spatial concentration of lost is readily apparent. In some communities, such as Greenwich Village and Brooklyn Heights, hardly any voters were disqualified from participating in the 2017 elections. In other communities, such as Harlem and Central Brooklyn, large numbers of individuals with a demonstrated history of voting were not allowed to cast a ballot for mayor.

@@

Matching methodologies, of course, only allow us to test the effect of being treated --- here, losing any voter for the 2017 election. The model above does not allow for different effects on turnout based on *how many* voters a neighborhood lost. In Table \ref{tab:trad-reg} below, I adopt a standard ordinary least squares regression to investigate whether lost voters are associated with lower turnout rates in the 2017 election. This regression uses the same covariates used in the matching procedure described above. Each neighborhood is weighted by its voting age population, and robust standard errors are clustered by city council district.^[Where neighborhoods cross city council district lines, they are assigned the district in which most of their voters live for clustering purporses.]
